---
title: "Lab 01"
author: "Amanda Mentesana"
date: "`2025-01-24`"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Reading the data set and filtering out NA values.
```{r}
EPI_data <- read.csv("C:/Users/amanda/Downloads/epi2024results06022024.csv") 

attach(EPI_data) #sets as default object, can call variables directly such as EPI_data$EPI.new is EPI.new

#head(EPI_data) #to show first part of data frame

# records True values if the value is NA
NAs <- is.na(EPI.new) 
# filters out NA values, new array 
EPI.new.noNAs <- EPI.new[!NAs]
```

## Excercise 1: Exploring the Distribution

```{r}
summary(EPI_data$EPI.new) #Output shows Median, Mean, Quartiles and range

fivenum(EPI.new,na.rm=TRUE) #Output shows the same values as summary() without heading 

stem(EPI.new) # stem and leaf plot 

hist(EPI.new) #basic histogram with default bin sizes

hist(EPI.new, seq(20., 80., 1.0), prob=TRUE) #histogram with a set range (20 to 80) and bin size of 1

lines(density(EPI.new,na.rm=TRUE,bw=1.)) # adds density plot over histogram, or try bw=“SJ” for a smoothed distribution

rug(EPI.new) #rug plot below graph 

boxplot(EPI.new, APO.new) #using a boxplot to visually compare variables for EPI.new and APO.new
```

The first histogram using the kernel density estimation shows there are two areas where the data may be centered, at about 42 and 65.
```{r}
#Histogram with kernal density estimation 

hist(EPI.new, seq(20., 80., 1.0), main=paste("Histogram with density line for EPI.new"), prob=TRUE) 

lines(density(EPI.new,na.rm=TRUE,bw="SJ")) 

rug(EPI.new)
```

To further look at these trends, using the normal distribution, we can see there are trends in the data which show two populations at 42 and 65.
```{r}
# histogram with normal distribution estimations
hist(EPI.new, seq(20., 80., 1.0, ),prob=TRUE) 

#lines(density(EPI.new,na.rm=TRUE,bw="SJ")) 

rug(EPI.new)

x<-seq(20,80,1) 

q<- dnorm(x,mean=42, sd=5,log=FALSE) #fitting to a normal curve distribution with mean centered at 42 and 5 standard deviations

lines(x,q) #plots line q over this range

lines(x,.4*q) # plots 40% of q over the range

p<-dnorm(x,mean=65, sd=5,log=FALSE) #fitting to a normal curve distribution with mean centered at 65 and 5 standard deviations

lines(x,.12*p) # plots 12% of p over the given range
```


## Excercise 2: Fitting a Distribution

Based on the distributions, the EPI.new variable reasonably fits the normal distribution fit, and shows possible bimodality.
```{r}
plot(ecdf(EPI.new), do.points=FALSE, verticals=TRUE) 

qqnorm(EPI.new) #a standard quantile-quantile plot of EPI dataset
qqline(EPI.new) #baseline

qqplot(rnorm(250), EPI.new, xlab = "Q-Q plot for norm dsn") #qq plot fit for the normal distribution
qqline(EPI.new)

qqplot(rt(250, df = 5), EPI.new, xlab = "Q-Q plot for t dsn") #qq plot plot fit to the t-dstribution with 5 degrees of freedom
qqline(EPI.new)


```


## Excercise 2a: Fitting the MPE Variable

We can see that the MPE variable has a different distribution and range than EPI. The mean is 69.85, yet both the 3rd quartile and maximum value is 100. When looking at the stem-and-leaf plot we can see that the data is skewed towards the value 100.
```{r}
summary(MPE.new) #Output shows Median, Mean, Quartiles and range

stem(MPE.new) # stem and leaf plot of SPI

boxplot(MPE.new, EPI.new) #using a box plot to visually compare variables for MPE.new and EPI.new

hist(MPE.new)
```

The first histogram using the kernal density estimation shows two peaks, with skewedness to 100.
```{r}
#Histogram with kernal density estimation 

hist(MPE.new, seq(0., 110., 10.), main=paste("Histogram with density line for MPE.new"), prob=TRUE) 

lines(density(MPE.new,na.rm=TRUE,bw="SJ")) 

rug(MPE.new)
```

To further look at these trends, using the normal distribution, we can see there are trends in the data which show two populations of data, one centered around 50 with a large spread of data and the other at 100 with a very narrow range of data.
```{r}
# histogram with normal distribution estimations
hist(MPE.new, seq(0., 110., 10., ),prob=TRUE) 

#lines(density(EPI.new,na.rm=TRUE,bw="SJ")) 

rug(MPE.new)

x<-seq(0,110,1) 

q<- dnorm(x,mean=50, sd=7,log=FALSE) #fitting to a normal curve distribution with mean centered at 50 and 7 standard deviations

lines(x,.3*q) # plots 0.3*q over the range

p<-dnorm(x,mean=95, sd=3,log=FALSE) #fitting to a normal curve distribution with mean centered at 95 and 3 standard deviations. The actual centering at 100 but using 95 is to fit to bin size for visulaization purposes..

lines(x,.6*p) # plots 0.6*p over the given range
```

The normal Q-Q plot shows that data is too peaked at the end of the distribution to fit it well.
```{r}
plot(ecdf(MPE.new), do.points=FALSE, verticals=TRUE) 

qqnorm(MPE.new); qqline(MPE.new) 

qqplot(rnorm(250), MPE.new, xlab = "Q-Q plot for norm dsn") 
qqline(MPE.new)

qqplot(rt(250, df = 5), MPE.new, xlab = "Q-Q plot for t dsn") 
qqline(MPE.new)
```


## Excercise 2b: Fitting the SPI Variable

We can see that the SPI variable has a larger spread than the EPI variable. The data also seems equally distributed using the stem and leaf plot.
```{r}
summary(SPI.new) #Output shows Median, Mean, Quartiles and range

#fivenum(SPI.new,na.rm=TRUE) #Output shows the same values as summary() without heading 

stem(SPI.new) # stem and leaf plot of SPI

boxplot(SPI.new, EPI.new) #using a boxplot to visually compare variables for SPI.new and EPI.new

hist(SPI.new)
```

The first histogram using the kernal density estimation shows very low variation in the SPI.new data distribution. Fitting to a normal curve was not helpful in describing the data.
```{r}
#Histogram with kernal density estimation 

hist(SPI.new, seq(0., 100., 10), main=paste("Histogram with density line for SPI.new"), prob=TRUE) 
lines(density(SPI.new,na.rm=TRUE,bw="SJ")) 
rug(SPI.new)

x<-seq(0,100,10) 

q<- dnorm(x,mean=50, sd=3,log=FALSE) #fitting to a normal curve distribution with mean centered at 42 and 5 standard deviations

#lines(x,q) #plots line q over this range
```

The normal Q-Q plot shows the SPI.new data is heavily tailed however it is fit to the data well.
```{r}

plot(ecdf(SPI.new), do.points=FALSE, verticals=TRUE) 

qqnorm(SPI.new)
qqline(SPI.new)  

#Make a Q-Q plot against the generating distribution by: 
qqplot(rnorm(250), SPI.new, xlab = "Q-Q plot for norm dsn") 
qqline(SPI.new)

qqplot(rt(250, df = 50), SPI.new, xlab = "Q-Q plot for t dsn") 
qqline(SPI.new)
```
